---
title: "Brent Crude Futures & Luxor Trading Strategy"
author: "Kopal Jha"
date: "10/2/2020"
output: pdf_document
---
This project was structured around financial data visualization and trading system design on R. Brent crude oil futures were selected for this analysis. Part 1 focuses on visualizing historical prices for Brent crude oil futures and summarizing important features of the historical distribution. Part 2 involves writing a script for the Luxor trading strategy from Jaekle and Tomasini's Trading Systems textbook. 

# Part 1 - Data Visualization
In the first part of this project, my objective was to get familiar with my dataset of Brent crude futures (historical prices) using baseR and the Quandl and ggplot packages for data visualization.

```{r}
library(Quandl)
library(tidyverse)
library(ggthemes)
library(quantmod)
library(quantstrat)
library(knitr)
library(blotter)

Quandl.api_key("httWUrzTVAL6RdqaHqSf")

brent <- Quandl('CHRIS/CME_BZ1') # calls first month contract for Brent crude futures on CME exchange
head(brent, n=10)

# WSJ-style plot of historical futures closing prices
ggplot(brent, aes(x = Date, y = Last)) +
  geom_line() +
  theme_wsj(color = "gray") +
  theme(axis.title=element_text(size=12)) +
  ggtitle("Brent Crude Futures Closing Prices") +
  theme(plot.title = element_text(size = 12, face = "bold"))
```
## Cleaning the data
```{r}
# Arrange data in reverse chronological order (oldest first) and sort by consecutive dates
brent <- brent %>% arrange(rev(rownames(.))) %>% mutate(date = as.Date(Date, "%d-%m-%Y")) %>% arrange(date)
brent$returns <- as.numeric(c('NA',diff(log(brent$Last)))) # Adding a new column with the differences between logs of Last prices
BZ_DATA <- select_if(brent, is.numeric)
# Get the number of NAN (not a number) values by column
na_cols_count <-sapply(BZ_DATA, function(y) sum(length(which(is.na(y)))))
na_cols_count[c("Open", "High", "Low", "Last", "Change", "Settle", "Volume", "Previous Day Open Interest", "returns")]
# This gives how many missing values are in each column. The "Change" column has too many missing values, so we select all columns except this one to incorporate into our working dataset.
BZ_DATA <- select(BZ_DATA,c(1,2,3,4,6,7,8,9))
# To account for the missing data in other columns, we select for only complete cases
BZ_DATA <- BZ_DATA[complete.cases(BZ_DATA),]
```

## Summary statistics
```{r}
summary(BZ_DATA) # Shortcut summary
(statistics <- do.call(data.frame, 
                      list(mean = round(apply(BZ_DATA, 2, mean),4),
                           sd = round(apply(BZ_DATA, 2, sd),4),
                           median = round(apply(BZ_DATA, 2, median),4),
                           min = round(apply(BZ_DATA, 2, min),4),
                           max = round(apply(BZ_DATA, 2, max),4)))) # Better formatted summary
```

## Distribution of returns
```{r}
returns <- brent$returns
qqnorm(returns, main="BZ Returns") 
qqline(returns, col="red") 
# The qqplots indicate the data is non-normally distributed
# The finding that the futures price data is non-normal is consistent with the fact that futures prices tend to follow a leptokurtic shape (fat tails, high peaks)
ggplot(BZ_DATA, aes(x = returns)) +
  geom_histogram() 
# The histogram confirms the leptokurtic nature of these prices
```

## Historical volatility
```{r}
dates <- brent$date
one_std <- sd(returns,na.rm = TRUE)
two_std <- one_std * 2 
plot(returns ~ dates, type='l',col='blue', main = "Returns over time")
abline(h=c(-one_std,one_std),col='red')
abline(h=c(-two_std,two_std),col='black')
# This plot shows that the Brent futures were most volatile in 2020.

plot(returns ~ dates, type='l',col='blue', main = "Returns over time (outliers removed)", ylim = c(-0.15, 0.15)) # Removing the -0.3 return still shows high volatility in 2020 compared to previous years
abline(h=c(-one_std,one_std),col='red')
abline(h=c(-two_std,two_std),col='black')

brent_volatility <- brent %>% subset(date > '2010-01-01') %>% mutate(year = substr(brent$date,1,4))
brent_volatility <- brent_volatility[complete.cases(brent_volatility), ]
# We then append a new column to this dataset using standard deviation as a measure of volatility
brent_volatility <- brent_volatility %>% group_by(year) %>% summarise(volatility = sd(returns)) %>% mutate(volatility)
ggplot(brent_volatility, aes(x=year, y=volatility)) +
  geom_col() +
  theme_wsj(color = "gray") +
  theme(axis.title=element_text(size=12)) +
  ggtitle("Volatility in Brent Crude Futures Returns (2014-2020)") +
  theme(plot.title = element_text(size = 12, face = "bold"))
```

## Candle charts and two technical indicators
```{r}
brent_vis <- Quandl('CHRIS/CME_BZ1', type = "xts")
candleChart(brent_vis)
addSMA(col="red")
addEMA()
colnames(brent_vis)[4] <- "Close"
brent_vis$date <- time(brent_vis) #Extracts dates as numeric values
```

## Bollinger Bands and volume
```{r}
chartSeries(brent_vis, subset="2014::2020",
            theme = chartTheme("white"),
            TA = "addVo(); addBBands(n = 20, sd = 2, draw = 'bands', on = -1)")
zoomChart("2020:", yrange=NULL) # Zooms in to show bands for 2020
```

## Autocorrelation of returns and closing prices
```{r}
acf(brent$returns, main = "Autocorrelation of returns", na.action = na.pass)
grid()
acf(brent$Last, main = "Autocorrelation of closing prices", na.action = na.pass)
grid()
# These autocorrelation plots indicate that closing prices are more autocorrelated than returns
```

# Part 2(a) - Luxor Trading Strategy
In the second part of this project, my objective was to script the Luxor trading strategy in order to better understand the workflow of trading strategy development. Luxor is a trend-following method that uses moving averages as entry signal generators. A long trade is initiated if the fast simple moving average (here, a 10-period SMA) is greater than or equal to the slow SMA (here, a 30-period SMA). If SMA(10) < SMA(30), a short order is submitted.

## Setup
```{r}
# Data
currency('USD')
sym <- na.omit(get(getSymbols("BZ=F"))["2019::2020"])
mkdata <- Cl(sym)
future("BZ=F", currency = "USD", multiplier = 1000)

# Set the parameters
initDate <- "2019-01-01" #Date of initiation 
from <- "2019-01-01" #Start date of the data set
to <- "2020-10-01" #End date of the data set
initEq <- 1e5 #Initial equity of $100,000

# Assign names to portfolio and account
strategy.st <- portfolio.st <- account.st <- "luxor"

# Initiate portfolio and account
initPortf(portfolio.st, "sym", initDate = initDate) #Initiate portfolio
initAcct(account.st, portfolios = strategy.st, initDate = initDate, initEq = initEq) #Initiate account
initOrders(portfolio = portfolio.st, initDate = initDate) #Initiate account
strategy(strategy.st, store = TRUE) #Store all the events in the strategy
```

## Indicators
```{r}
add.indicator(strategy = strategy.st,
              name = "SMA",
              arguments = list(x = quote(Cl(mkdata)), # Note these indicators are using closing prices to calculate the moving averages
                               n = 10),
              label = "nFast") # Fast Moving Average ie SMA(10)

add.indicator(strategy = strategy.st,
              name = "SMA",
              arguments = list(x = quote(Cl(mkdata)),
                               n = 30),
              label = "nSlow") # Slow Moving Average ie SMA(30)
```

## Signals
```{r}
add.signal(strategy = strategy.st,
           name="sigCrossover",
           arguments = list(columns = c("nFast", "nSlow"),
                            relationship = "gte"), # The long signal occurs when the fast SMA is > or = the slow SMA
           label = "long")

add.signal(strategy = strategy.st,
           name="sigCrossover",
           arguments = list(columns = c("nFast", "nSlow"),
                            relationship = "lt"), # The short signal occurs when the fast SMA is < the slow SMA
           label = "short")
```

## Entry rules
```{r}
add.rule(strategy = strategy.st,
         name = "ruleSignal",
         arguments = list(sigcol = "long",
                          sigval = TRUE,
                          orderqty = 100,
                          ordertype = "stoplimit",
                          orderside = "long", 
                          threshold = 0.005,
                          prefer = "High", 
                          TxnFees = -10,  # Assuming my broker charges $10 for placing a long order
                          replace = FALSE), # Will not replace any open orders
         type = "enter",
         label = "EnterLONG")

add.rule(strategy.st,
         name = "ruleSignal",
         arguments = list(sigcol = "short",
                          sigval = TRUE,
                          orderqty = -100,
                          ordertype = "stoplimit",
                          threshold = -0.005, 
                          orderside = "short", 
                          replace = FALSE, 
                          TxnFees = -10, 
                          prefer = "Low"),
         type = "enter",
         label = "EnterSHORT")
```

## Exit rules
```{r}
add.rule(strategy.st,
         name = "ruleSignal",
         arguments = list(sigcol = "short",
                          sigval = TRUE,
                          orderside = "long", # We exit our long positions when going short
                          ordertype = "market", # Buys the future at the prevailing market price
                          orderqty = "all",
                          TxnFees = -10,
                          replace = TRUE), # Any open orders are replaced
         type = "exit",
         label = "Exit2SHORT")

add.rule(strategy.st,
         name = "ruleSignal",
         arguments = list(sigcol = "long",
                          sigval = TRUE,
                          orderside = "short",
                          ordertype = "market",
                          orderqty = "all",
                          TxnFees = -10,
                          replace = TRUE),
         type = "exit",
         label = "Exit2LONG")
```

## Execute
```{r}
# Apply the strategy
applyStrategy(strategy.st, portfolio.st)

# Update portfolio, account, and equity
updatePortf(portfolio.st)
updateAcct(account.st)
updateEndEq(account.st)
```

# Part 2(b) - Performance Analysis
In part 2(b), I analyzed the results of my trading strategy using several performance measures.

## Simple plot of performance
```{r}
myTheme <- chart_theme()
myTheme$col$dn.col<-'lightblue'
myTheme$col$dn.border <- 'lightgray'
myTheme$col$up.border <- 'lightgray'
chart.Posn(strategy.st,theme=myTheme,
           TA='add_SMA(n=10,col=4, on=1, lwd=2)')
```

## Trade Statistics
```{r}
tstats <- tradeStats(portfolio.st)
kable(t(tstats))
```

## Trade Related Statistics
Segments tradeStats output to more cleanly display metrics related to trades only.
```{r}
require(dplyr)
tab.trades <- tstats %>% 
  mutate(Trades = Num.Trades, 
         Win.Percent = Percent.Positive, 
         Loss.Percent = Percent.Negative, 
         WL.Ratio = Percent.Positive/Percent.Negative) %>% 
  select(Trades, Win.Percent, Loss.Percent, WL.Ratio)
kable(t(tab.trades))
```

## Profit Related Statistics
Segments tradeStats output to more cleanly display metrics related to profit only.
```{r}
tab.profit <- tstats %>% 
  select(Net.Trading.PL, Gross.Profits, Gross.Losses, Profit.Factor)
kable(t(tab.profit))
```

## Averages
Segments tradeStats output to more cleanly display metrics related to averages only.
```{r}
tab.wins <- tstats %>% 
  select(Avg.Trade.PL, Avg.Win.Trade, Avg.Losing.Trade, Avg.WinLoss.Ratio)

kable(t(tab.wins))
```

## Performance Summary
```{r}
rets <- PortfReturns(Account = account.st)
rownames(rets) <- NULL
charts.PerformanceSummary(rets, colorset = bluefocus)
```

## Performance Statistics
```{r}
tab.perf <- table.Arbitrary(rets,
                            metrics=c(
                              "Return.cumulative",
                              "Return.annualized",
                              "SharpeRatio.annualized",
                              "CalmarRatio"),
                            metricsNames=c(
                              "Cumulative Return",
                              "Annualized Return",
                              "Annualized Sharpe Ratio",
                              "Calmar Ratio"))
kable(tab.perf)
```

## Risk Statistics
```{r}
tab.risk <- table.Arbitrary(rets,
                            metrics=c(
                              "StdDev.annualized",
                              "maxDrawdown",
                              "VaR",
                              "ES"),
                            metricsNames=c(
                              "Annualized StdDev",
                              "Max DrawDown",
                              "Value-at-Risk",
                              "Conditional VaR"))
kable(tab.risk)
```

## Order book
```{r}
(ob <- getOrderBook(portfolio.st))
```

## Account summary
```{r}
require(lattice)
a <- getAccount(account.st)
xyplot(a$summary, type = "h", col = 4)
```

## Equity curve
```{r}
equity <- a$summary$End.Eq
plot(equity, main = "Equity Curve")
```

# Future Directions
This project enabled me to strengthen my familiarity with trading strategy scripting in R. The Luxor strategy was selected because of its simplicity and popularity among quantstrat tutorials. Future projects may focus on selecting strategies that target the specific features of Brent crude oil futures. One interesting project may involve testing a binary options trading strategy for BZ futures to capture volatility around inventory releases, with signals for inventory releases specified through the sigTimestamp argument.